The Evolution of Phenotypically Invariant Genetic Networks

### 2017-10-21 kalman_walk_paper.tex
rewrote apx about limiting OU process

\section{Genetic drift with a multivariate trait}
\label{ss:quant_gen}

For completeness, we provide a brief argument of how the population mean
moves under genetic drift
with a quantitative genetics model,
as in \citet{lande1981models} or \citet{hansen1996translating}.
These ignore details of the underlying genetic basis,
but developing a more accurate model is beyond the scope of this paper.

\paragraph{Completing the square}
First note that 
\begin{align*}
    (x-y)^T A (x-y)
    &=
    x^T A \left( x - 2y \right) + y^T A y ,
\end{align*}
and so
\begin{align*}
    (x-y)^T A (x-y) + x^T B x
    &=
    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
    &=
    \left( x - (A + B)^{-1} A y \right)^T
    (A + B)
    \left( x - (A + B)^{-1} A y \right)
    + \text{(terms that don't depend on $x$)} .
\end{align*}
Therefore, if $f(x;\Sigma,y)$ is the density of a Gaussian with mean $y$ and covariance matrix $\Sigma$
then substituting $A=\Sigma^{-1}$ and $B=U^{-1}$ above,
\begin{align*}
    \frac{ f(x;\Sigma,y) f(x;U,0) }{\int_x f(z;\Sigma,y) f(z;U,0) dz}
    &=
    f(x; (\Sigma^{-1} + U^{-1})^{-1}, (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} y) .
\end{align*}

Now suppose that the population is distributed in genotype space
as a Gaussian with covariance matrix $\Sigma$ and mean $y$.
Selection has the effect of multiplying this density by the fitness function and renormalizing,
so that if expected fitness of $x$ is proportional to $f(x;U,z)$
then the above argument shows that the next generation will be sampled from a Gaussian distribution
with covariance matrix $(\Sigma^{-1} + U^{-1})^{-1}$ 
and mean $z + (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} (y-z)$
Taking a sample of size $N$ to construct the next generation 
will produce something close to this but with a slightly (stochastically) deviating mean.
The next generation's mean is drawn from a Gaussian distribution with mean
with covariance matrix $(\Sigma^{-1} + U^{-1})^{-1}/N$ 
and mean $z + (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} (y-z)$.

Roughly, what is this doing?
Suppose that the population mean differs from the optimum by $\epsilon$,
that $\Sigma = \sigma^2 I$ and $U = I/\beta^2$ (so, stablizing selection happening on a distance scale of $\beta$).
Then the population mean gets closer to the optimum on average, moving to
$\epsilon/(1 + \sigma^2 \beta^2)$
and adds noise of size $(1/\beta) \sigma/\sqrt{N \sigma^2 + N 1/\beta^2}$.
At equilibrium, these two movements will be of the same order,
so that $\epsilon$ is of order $(\sigma/\sqrt{N}) \sqrt{1+\sigma^2 \beta^2}$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Genetic drift with a multivariate trait}
\label{ss:quant_gen}

For completeness, we provide a brief exposition of how a population 
evolves due to genetic drift
with a quantitative genetics model,
as in \citet{lande1981models} or \citet{hansen1996translating}.
These ignore details of the underlying genetic basis,
but developing a more accurate model is beyond the scope of this paper.

Suppose that the population is distributed in trait space
as a Gaussian with covariance matrix $\Sigma$ and mean $\mu$,
whose density we write as $f(\cdot;\Sigma,\mu)$.
Selection has the effect of multiplying this density by the fitness function and renormalizing,
so that if expected fitness of $x$ is proportional to $f(x;U,0)$,
then the distribution post-selection
has density at $x$ proportional to $f(x;\Sigma,\mu) f(x;U,0)$.
By the computation below (``Completing the square''),
the result is a Gaussian distribution
with covariance matrix $(\Sigma^{-1} + U^{-1})^{-1}$ 
and mean $(\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} \mu$.
Importantly, if $U$ is not invertible, then by $U^{-1}$ 
we mean $(U^T U)^{-1} U^T$, Moore-Penrose pseudoinverse of $U$,
so that $U U^{-1}$ is equal to the projection matrix
onto the span of the columns of $U$.

After selection, we have reproduction:
suppose this occurs as in the infinitesimal model \citep{infinitesimal},
so that each offspring of parents with traits $x$ and $y$
is drawn independently from a Gaussian distribution 
with mean $(x+y)/2$ and covariance matrix $N$.
Here, $N$ is the contribution of ``segregation variance''.
If $\widetilde \Sigma = (\Sigma^{-1} + U^{-1})^{-1}$ 
is the covariance matrix of the parents post-selection,
then the distribution of offspring will again be Gaussian,
with mean equal to that of the parents
and covariance matrix $\widetilde \Sigma/2 + N$.

In summary, a generation under this model modifies the mean ($\mu$)
and covariance matrix ($\Sigma$) of a population as follows:
\begin{align*}
    \mu &\mapsto \mu' = (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} \mu \\
    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + U^{-1})^{-1} + N .
\end{align*}
What measures are stable under this transformation?
The condition $\mu = \mu'$ reduces to $\Sigma U^{-1} \mu = 0$;
if we assume $N$ and therefore $\Sigma$ are of full rank,
then this happens if and only if $\mu$ is not in the span of the columns of $U$,
i.e., if $\mu$ lies in a neutral direction.
The condition $\Sigma' = \Sigma$ can also be solved,
at least numerically.
After rearrangement, it reduces to
$\Sigma U^{-1} \Sigma + (I/2 - N U^{-1}) \Sigma = N$.
We can find a more explicit description
if we assume that $x^T U x = \sum_{i=1}^k x_i^2$,
i.e., that selection only cares about the first $k$ coordinates,
and then with no interactions between traits.
If so, the condition $\Sigma' = \Sigma$
can be written in block form as
\begin{align*}
    \begin{bmatrix}
        \Sigma_{11}^2 + (I/2 - N_{11}) \Sigma_{11}
        & \Sigma_{11} \Sigma_{12} + (I/2 - N_{11})\Sigma_{12} \\
        \Sigma_{12}^T \Sigma_{11} + \Sigma_{12}^T/2 - N_{12}^T \Sigma_{11}
        & \Sigma_{22} - N_{12}^T \Sigma_{12} 
    \end{bmatrix}
    &=
    \begin{bmatrix}
        N_{11} & N_{12} \\
        N_{12}^T & N_{22}
    \end{bmatrix} .
\end{align*}
The first equation, $\Sigma_{11}^2 + (I/2 - N_{11}) \Sigma_{11} = N_{11}$,
can be solved with the quadratic formula: 
$$\Sigma_{11} = (N_{11} - I/2 + Q)/2$$
for any $Q$ that commutes with $N_{11}$
and is a solution to $Q^2 = (N_{11} - I/2)^2 + 4 N_{11}$.
Since we need $\Sigma$ to be positive definite,
we take the solution with positive eigenvalues.
Given $\Sigma_{11}$, the remaining components are
\begin{align*}
    \Sigma_{12} 
    &= 
    \left( \Sigma_{11} + I/2 - N_{11} \right)^{-1} N_{12} \\
    \Sigma_{22}
    &=
    N_{12}^T \Sigma_{12} + N_{22} .
\end{align*}
Importantly, the mean $\mu$ does not affect 
either how the covariance matrix moves,
or its stable shape.

Above we have described the \emph{expected} motion of the mean and covariance..
However, random resampling will introduce noise.
Suppose that a population of $N$ individuals
behaves approximately as described above.
By the above,
we may expect that the covariance matrix stays close to a constant value $\Sigma$,
computed from $N$ and $U$ as above,
so that we need only consider motion of the mean, $\mu$.
Since we take a sample of size $N$ to construct the next generation,
the next generation's mean is drawn from a Gaussian distribution with mean $\mu'$
and covariance matrix $\Sigma/N$.
Defining $\Gamma = (I - (I + \Sigma U^{-1})^{-1})$,
this can be written as
\begin{align*}
    \mu' - \mu = \Gamma \mu + \epsilon/\sqrt{N} ,
\end{align*}
where $\epsilon$ is a multivariate Gaussian with mean zero and covariance matrix $\Sigma$.
Let $\mu(k)$ denote the mean in the $k^\text{th}$ generation,
and suppose that $\mu$ differs from optimal by something of order $1/N$:
if $\nu(t) = N \mu(tN)$,
then the previous equation implies that as $N \to \infty$, 
in the limit $\nu$ solves the It\^{o} equation
\begin{align*}
    d \nu(t) = \Gamma \nu(t) dt + \Sigma^{1/2} dW(t) ,
\end{align*}
where now $W(t)$ is a multivariate white noise.
This has an explicit solution as a multivariate Ornstein-Uhlenbeck process:
\begin{align*}
    \nu(t) = e^{-t \Gamma} \nu_0 + \int_0^t e^{-(t-s) \Gamma} \Sigma^{1/2} dW(s).
\end{align*}
% Note that after some algebra, $\Gamma = \Sigma U^{-1} (I - \Sigma U^{-1})^{-1}$.
The asymptotic variance of this process in the direction $z$ is
\begin{align} \label{eqn:limiting_cov}
    \lim_{t \to \infty} \var[\nu(t) \cdot z]
    &=
    \int_0^\infty z^T e^{-s \Gamma} \Sigma e^{-s \Gamma} z ds ,
\end{align}
which is infinite iff $\Gamma z = 0$,
which occurs iff $Uz=0$.
In other words, population mean trait values lie away from the optimal set
by a Gaussian displacement of order $1/N$
with a covariance matrix given by equation \eqref{eqn:limiting_cov}.

\plri{Now write what this means intuitively.}

\paragraph{Completing the square}
First note that 
\begin{align*}
    (x-y)^T A (x-y)
    &=
    x^T A \left( x - 2y \right) + y^T A y ,
\end{align*}
and so
\begin{align*}
    (x-y)^T A (x-y) + x^T B x
    &=
    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
    &=
    \left( x - (A + B)^{-1} A y \right)^T
    (A + B)
    \left( x - (A + B)^{-1} A y \right) \\
    &\qquad{}
    + \text{(terms that don't depend on $x$)} .
\end{align*}
Therefore, 
by substituting $A=\Sigma^{-1}$ and $B=U^{-1}$,
\begin{align*}
    \frac{ f(x;\Sigma,y) f(x;U,0) }{\int f(z;\Sigma,y) f(z;U,0) dz}
    &=
    f(x; (\Sigma^{-1} + U^{-1})^{-1}, (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} y) .
\end{align*}
